Gusto: a toolkit for compatible finite element dynamical cores

### 2024-08-09 source/paper.tex
some edits

The finite element method represents fields on a mesh made of
non-overlapping, conforming cells, or elements. Within each element,
the field is represented as a sum of coefficients multiplied by
polynomial basis functions. The finite element space is specified by
the degree of the polynomial basis functions and the continuity of
these functions between the elements. The partial differential
equations used in weather and climate modelling have multiple
prognostic fields and it is not necessary, or indeed desirable, to use
the same finite element function space for each field. Using different
finite element function spaces for each field gives what is known as a
mixed finite element discretisation. If these different finite element
function spaces are chosen so that discrete versions of the vector
calculus identities ** hold then the discretisation is called
`mimetic' or `compatible'.

In this article we will describe the current capabilities of Gusto and
the design features that enable our flexible implementation whereby
different spatial and temporal discretisations can be applied to a
range of different equation sets relevant for weather and climate
modelling. In section \ref{sec: governing} we will outline three
different equation sets, along with their finite element
discretisations. Then in section \ref{sec: design} we will present the
code structure of Gusto followed by a detailed example (section
\ref{sec: FML}) of the Form Manipulation Language (FML) that
facilitates symbolic manipulation of the finite element forms within
our timestepping code. Section \ref{sec: results} contains a selection
of numerical results that showcases the range of discretisations
available within Gusto. **mention here new functionality facilitated
by the flexibility of Gusto? We summarise our current status and future
directions in section \ref{sec: summary}.

@@

\subsection{Compressible Boussinesq Equations}
The compressible Boussinesq equations are commonly used in ocean
modelling. They are useful from a numerical discretisation perspective
because they incorporate compressibility effects without the
complexity of discretising the nonlinear pressure gradient terms that
we shall see in the compressible Euler equations. We write these
equations in terms of the prognostic variables $\MM{u}$, $b$ and $p$
where $\MM{u}$ is a three dimensional velocity field, $b$ is the fluid
buoyancy and $p$ is pressure. In terms of these variables, the
equations are:

\begin{align}
  \pp{\MM{u}}{t} + 
  (\MM{u}\cdot\nabla)\MM{u} +
  2\MM{\Omega}\times \MM{u} + \nabla p + b\hat{\MM{k}} & = 0, \\
  \pp{p}{t} + c^2\nabla\cdot\MM{u} & = 0, \\
  \pp{b}{t} + (\MM{u}\cdot\nabla) b & = 0,
\end{align}
where $\MM{\Omega}$ is the rotation vector of the domain,
$\hat{\MM{k}}$ is the unit vector pointing upwards and $c$ is the
acoustic wave speed. The velocity $\MM{u}$ can be two or three
dimensional, depending on whether the domain is a vertical slice
described by $(x, z)$ coordinates, of a three dimensional domain
described by $(x, y, z)$ coordinates.

@@

\begin{align}
  \pp{\MM{u}}{t} + 
  (\MM{u}\cdot\nabla)\MM{u} +
  2\MM{\Omega}\times \MM{u} + c_p\theta\nabla \Pi + g\hat{\MM{k}} & = 0, \\
  \pp{\rho}{t} + \nabla\cdot(\MM{u}\rho) & = 0, \\
  \pp{\theta}{t} + (\MM{u}\cdot\nabla)\theta & = 0, \\
  \Pi^{(1-\kappa)/\kappa} & = \frac{R}{p_0}\rho\theta, & 
\end{align}
where $\Pi$ is the Exner pressure, $p_0$ is a reference pressure,
typically taken to be *, $R$ is ** and $\kappa$ is **. The velocity
$\MM{u}$ can be two or three dimensional, depending on whether the
domain is a vertical slice described by $(x, z)$ coordinates, of a
three dimensional domain described by $(x, y, z)$ coordinates.

@@

Comparing the equation sets given in the previous sections, we see
several similarities. For example, the prognostic equation for the
velocity always contains a nonlinear transport term, a term due to
rotation and a `pressure' gradient term (in shallow water this is the
gradient of the depth but the term plays the same role in the dynamics
and is derived from the pressure gradient term in the full equations
\citep{zeitlin2018geophysical}). The other prognostic fields satisfy
either an advective or continuity form of the transport
equation. These similarities remain when we formulate the compatible
finite element discretisations, as we shall see below, and importantly
this means that the discretisation of these terms can be written in a
general way that means it can be applied to any of the equation sets.

@@

In line with the theory outlined in section \ref{}, we always choose
the velocity $\MM{u} \in \mathbb{V}_u$ with $\mathbb{V}_u \subset
H(\text{div})$. The field whose gradient appears in the `pressure'
gradient term is chosen to be in the space $\mathbb{V}_p$ with
$\mathbb{V}_p \subset L_2$. The compressible Boussinesq and
compressible Euler equations have a third prognostic variable, either
buoyancy or potential temperature, which is chosen to be in the space
$\mathbb{V}_\theta$ which has degrees of freedom that are collocated
with those of the vertical velocity, mimicking the Charney-Phillips
staggering as discussed in **. These choices are governed by the
properties required of discretisation of the linearised equation sets
(as discussed in section ** and other refs). To solve the nonlinear
equation sets we need to introduce stable and accurate discretisations
of the transport terms and the nonlinear pressure gradient term in the
compressible Euler equation set. In addition, some test cases require
the addition of viscous and diffusive terms to generate converged
solutions \ref{}. In the following subsections we illustrate a
non-exhaustive (**probably!) selection of the options available in
Gusto.

@@

\begin{equation}
\int_\Omega\MM{w}\cdot\nabla p \diff x = -\int_\Omega\nabla\cdot\MM{w}p\diff x,
\end{equation}
where $\Omega$ denotes the domain and there are no surface terms from
the integration by parts because these vanish on interior element
boundaries since the normal component of $\MM{w}$ is continuous across
element facets and on exterior boundaries we set no normal flow
boundary conditions. However, the pressure gradient term in the
compressible Euler equations, $c_p\theta\nabla\Pi$, is nonlinear and
this introduces an additional complication. Integrating by parts is
still required since $\Pi$ is evaluated as a function of the density
$\rho \in L_2$ and so is discontinuous, but this means that we now
have a differential operator applied to $\theta \in \mathbb{V}_\theta$
which is discontinuous in the vertical. Following \ref{} we integrate
by parts in each element $e$, giving

@@

The finite element discretisation of each term in an equation is
represented in Firedrake by a \texttt{ufl.Form} object. FML defines a
Label object which tags a ufl.Form to become a Term, with the
intention that the Term represents a term in the equation.

FML defines a
\texttt{Term} object that comprises the \texttt{ufl.Form} defining the
finite element discretisation of the term and a dictionary of
key-value pairs corresponding to the FML \texttt{Label} objects that
apply to this term. An FML \texttt{Label} object has a name and an
optional value, defaulting to \texttt{True} to indicate the presence
of the label. Labels can also be used to tag parts of the form. For
example, the \texttt{subject} label is used to tag the
\texttt{Function} in the \texttt{ufl.Form} that represents the
prognostic variable. These operations are composable, so more than one
label can be attached to a term. The end result of labelling the terms
in the residual is a \texttt{LabelledForm} object that holds a list of
terms, pairing individual \texttt{ufl.Form} objects with FML
labels. Labelled forms can be added together, multiplied by floats and
have additional terms added. More complex manipulations are enabled by
the \texttt{label\_map} method that provides a way of filtering the
terms in the form based on their labels, applying different mappings
to terms that pass or fail the conditions defined in the filter. A
common operation might be to filter for terms that have a certain
label, such as \texttt{transport}, drop terms that do not have this
label and multiply those that do by the timestep. We shall demonstrate
this in the following example.

@@

We demonstrate the use of FML by applying a split timestepping scheme
to the 1D advection diffusion equation
\begin{equation}\label{eq:1dadvdiff}
  q_t + cq_x - \kappa q_{xx} = 0,
\end{equation}
where $q$ is a scalar field, $c$ is the constant transporting velocity
and $\kappa$ is the constant diffusivity. The time discretisation
applies an explicit step of forward Euler to the transport term and
an implicit step of backward Euler to the diffusion term:
\begin{align}
  q^* &= q^n - c \Delta t q_x^n, \label{eq:explicit-advdiff} \\
  q^{n+1} - \kappa \Delta t q_{xx}^{n+1} &= q^*, \label{eq:implicit-advdiff}
\end{align}
where $q^n$ is the value of $q$ at time step $n$, $q^*$ is an
intermediate value for $q$ and $\Delta t$ is the timestep.

We illustrate the use of FML for this example in three steps: 1)
define and label the weak form of the residual; 2) build solvers for
the explicit and implicit systems; 3) run the timestepping loop. Since
the intention of this example is to demonstrate FML rather than the
compatible finite element discretisation, we use space of piecewise
quadratic continuous Lagrange polynomials, denoted by $P_2$, for the
finite element discretisation as this greatly simplifies the
forms. The finite element discretisation of equation
\ref{eq:1dadvdiff} is then
\begin{equation}\label{eq:weak-advdiff}
  \int_\Omega vq_t \diff x + c \int_\Omega v q_x \diff x + \kappa \int_\Omega v_x q_x \diff x = 0, \quad \forall v \in P_2,
\end{equation}
where $v\in P_2$ is the test function and the diffusion term has been
integrated by parts.

@@

\lstinputlisting[caption={This code defines the solver for the explicit step of the timestepping method i.e. equation \ref{eq:explicit-advdiff}. \texttt{qn} and \texttt{qstar} are Firedrake \texttt{Functions} with \texttt{qn} holding the values of $q$ at time step $n$ and \texttt{qstar} holding the values after the explicit solve.}, label={lst:define-explicit}, language=python, firstline=40, lastline=55]{code/advection_diffusion_splitting_example.py}

The implicit solver is defined in a similar way (see listing
\ref{lst:define-implicit}). To create the left side of equation
\ref{eq:implicit-advdiff} we begin with the original residual, drop
the transport term (line 1), multiply all remaining terms that are not
the time derivative by $\Delta t$ (line 3) and replace the subject
with a trail function (line 5). The right side only contains the form
coming from the discretisation of the time derivative (line 7),
evaluated using the result of the explicit solve, \texttt{qstar} (line
9).

@@

\lstinputlisting[caption=blah, label={lst:timeloop}, language=python, firstline=73, lastline=79]{code/advection_diffusion_splitting_example.py}

>>>>

The finite element method represents fields (e.g. velocity, pressure,
density or temperature) on a mesh made of non-overlapping, conforming
cells, or elements. Within each element, the field is represented as a
sum of coefficients multiplied by polynomial basis functions. The
finite element space is specified by the degree of the polynomial
basis functions and the continuity of these functions between the
elements. The partial differential equations used in weather and
climate modelling have multiple prognostic fields and it is not
necessary, or indeed desirable, to use the same finite element
function space for each field. Using different finite element function
spaces for each field gives what is known as a mixed finite element
discretisation. If these different finite element function spaces are
chosen so that discrete versions of vector calculus identities hold
then the discretisation is called `mimetic' or `compatible'.

In this article we will describe the current capabilities of Gusto and
the design features that enable our flexible implementation whereby
different spatial and temporal discretisations can be applied to a
range of different equation sets relevant for weather and climate
modelling. In section \ref{sec: governing} we will outline three
different equations commonly used for modelling geophysical fluids in
order to demonstrate the similaries between the equations and justify
the use of a unified codebase. The advantages of this unified approach
become even clearer in section \ref{sec: FEM} where we present the
compatible finite element discretisation of the different types of
terms (e.g. pressure gradient, transport) seen in section \ref{sec:
  governing}. In section \ref{sec: design} we describe the code
structure of Gusto followed by a detailed example (section \ref{sec:
  FML}) of the Form Manipulation Language (FML) that facilitates
symbolic manipulation of the finite element forms within our
timestepping code. Section \ref{sec: results} contains a selection of
numerical results that showcases the range of discretisations
available within Gusto. **mention here new functionality facilitated
by the flexibility of Gusto? We summarise our current status and
future directions in section \ref{sec: summary}.

@@

\subsection{Compressible Boussinesq Equations}
The compressible Boussinesq equations are commonly used in ocean
modelling. They are useful from a numerical discretisation perspective
because they incorporate compressibility effects without the
complexity of discretising the nonlinear pressure gradient terms that
we shall see in the compressible Euler equations. We write these
equations in terms of the prognostic variables $\MM{u}$, $b$ and $p$
where $\MM{u}$ is the velocity field, $b$ is the fluid buoyancy and
$p$ is pressure. In terms of these variables, the equations are:

\begin{align}
  \pp{\MM{u}}{t} + 
  (\MM{u}\cdot\nabla)\MM{u} +
  2\MM{\Omega}\times \MM{u} + \nabla p + b\hat{\MM{k}} & = 0, \\
  \pp{p}{t} + c^2\nabla\cdot\MM{u} & = 0, \\
  \pp{b}{t} + (\MM{u}\cdot\nabla) b & = 0,
\end{align}
where $\MM{\Omega}$ is the rotation vector of the domain,
$\hat{\MM{k}}$ is the unit vector pointing upwards and $c$ is the
acoustic wave speed. The equations have been written in
three-dimensional form but they can also be solved in a
two-dimensional vertical slice domain described by $(x, z)$
coordinates. In this case the velocity $\MM{u}$ can either be
two-dimensional as well (if there is no Coriolis force), or fully
three-dimensional but with each cartensian component only varying in
the $x-z$ plane. In the latter case this is implemented using a
three-dimensional computational domain that is only one element wide
in the $y$-direction and with periodic boundary conditions also in
that direction.

@@

\begin{align}
  \pp{\MM{u}}{t} + 
  (\MM{u}\cdot\nabla)\MM{u} +
  2\MM{\Omega}\times \MM{u} + c_p\theta\nabla \Pi + g\hat{\MM{k}} & = 0, \\
  \pp{\rho}{t} + \nabla\cdot(\MM{u}\rho) & = 0, \\
  \pp{\theta}{t} + (\MM{u}\cdot\nabla)\theta & = 0, \\
  \Pi^{(1-\kappa)/\kappa} & = \frac{R}{p_0}\rho\theta, & 
\end{align}
where $\Pi$ is the Exner pressure, $p_0$ is a reference pressure, $R$
is the gas constant, $c_p$ is the specific heat at constant pressure
and $\kappa=R/c_p$. The equations have been written in
three-dimensional form but they can also be solved in a
two-dimensional vertical slice domain described by $(x, z)$
coordinates. In this case the velocity $\MM{u}$ can either be two
dimensional as well (if there is no Coriolis force), or fully three
dimensional but with each cartensian component only varying in the
$x-z$ plane. In the latter case this is implemented using a three
dimensional computational domain that is only one element wide in the
$y$-direction and with periodic boundary conditions also in that
direction. \JScomment{I know this repeats what is said above but I
  like that each section is self-contained - I hope that is ok but
  also happy to change it!}

@@

Comparing the equation sets given in the previous sections, we see
several similarities. For example, the prognostic equation for the
velocity always contains a nonlinear transport term, a term due to
rotation and a `pressure' gradient term (in shallow water this is the
gradient of the depth but the term plays the same role in the dynamics
and is derived from the pressure gradient term in the full equations
\citep{zeitlin2018geophysical}). The other prognostic fields satisfy
either an advective or continuity form of the transport
equation. These similarities remain when we formulate the compatible
finite element discretisations, as we shall see below, and importantly
this means that the discretisation of these terms can be written and
implemented in software so that it can be applied to any of the
equation sets.

@@

In line with the theory outlined in section \ref{}, we always choose
the velocity $\MM{u} \in \mathbb{V}_u$ with $\mathbb{V}_u \subset
H(\text{div})$. The field whose gradient appears in the `pressure'
gradient term is chosen to be in the space $\mathbb{V}_p$ with
$\mathbb{V}_p \subset L_2$, or, in the case of the more complex
nonlinear pressure gradient term for the compressible Euler equations,
it is a function of a variable in this space. The compressible
Boussinesq and compressible Euler equations have a third prognostic
variable, either buoyancy or potential temperature, which is chosen to
be in the space $\mathbb{V}_\theta$ which has degrees of freedom that
are collocated with those of the vertical velocity, mimicking the
Charney-Phillips staggering as discussed in **. These choices are
governed by the properties required of discretisation of the
linearised equation sets (as discussed in section ** and other
refs). To solve the nonlinear equation sets we need to introduce
stable and accurate discretisations of the transport terms and the
nonlinear pressure gradient term in the compressible Euler equation
set. In addition, some test cases require viscous and diffusive terms
to generate converged solutions \ref{}. In the following subsections
we illustrate a non-exhaustive (**probably!) selection of the options
available in Gusto.

@@

\begin{equation}
\int_\Omega\MM{w}\cdot\nabla p \diff x = -\int_\Omega\nabla\cdot\MM{w}p\diff x, \quad \forall \MM{w} \in \mathbb{V}_u,
\end{equation}
where $\Omega$ denotes the domain and there are no surface terms from
the integration by parts because these vanish on interior element
boundaries since the normal component of $\MM{w}$ is continuous across
element facets and on exterior boundaries we set no normal flow
boundary conditions. However, the pressure gradient term in the
compressible Euler equations, $c_p\theta\nabla\Pi$, is nonlinear and
this introduces an additional complication. Integration by parts is
still required since $\Pi$ is evaluated as a function of the density
$\rho \in \mathbb{V}_\rho$ and $\theta \in \mathbb{V}_\theta$ so is
discontinuous, but this means that we now have a differential operator
applied to $\theta \in \mathbb{V}_\theta$ which is discontinuous in
the vertical. Following \ref{} we integrate by parts in each element
$e$, giving

@@

The finite element discretisation of each term in an equation is
represented in Firedrake by a UFL form object. FML defines a
\texttt{Label} object which is used to tag a UFL form to create a
\texttt{Term}, with the intention that the \texttt{Term} represents a
term in the residual.

FML defines a \texttt{Term} object that comprises the
\texttt{ufl.Form} defining the finite element discretisation of the
term and a dictionary of key-value pairs corresponding to the FML
\texttt{Label} objects that apply to this term. An FML \texttt{Label}
object has a name and an optional value, defaulting to \texttt{True}
to indicate the presence of the label. Labels can also be used to tag
parts of the form. For example, the \texttt{subject} label is used to
tag the \texttt{Function} in the \texttt{ufl.Form} that represents the
prognostic variable, or `subject' of the equation. These operations
are composable, so more than one label can be attached to a term. The
end result of labelling the terms in the residual is a
\texttt{LabelledForm} object that holds a list of terms, pairing
individual \texttt{ufl.Form} objects with FML labels. Labelled forms
can be added together, multiplied by floats and have additional terms
added. More complex manipulations are enabled by the
\texttt{label\_map} method that provides a way of filtering the terms
in the form based on their labels, applying different mappings to
terms that pass or fail the conditions defined in the filter. A common
operation might be to filter for terms that have a certain label, such
as \texttt{transport}, drop terms that do not have this label and
multiply those that do by the timestep. We shall demonstrate this in
the following example.

@@

We demonstrate the use of FML by applying a split timestepping scheme
to the 1D advection diffusion equation
\begin{equation}\label{eq:1dadvdiff}
  q_t + cq_x - \kappa q_{xx} = 0,
\end{equation}
where $q$ is a scalar field, $c$ is the constant transporting velocity
and $\kappa$ is the constant diffusivity. The time discretisation
applies an explicit timestep of forward Euler to the transport term
$cq_x$ and an implicit timestep of backward Euler to the diffusion
term $\kappa q_{xx}$:
\begin{align}
  q^* &= q^n - c \Delta t q_x^n, \label{eq:explicit-advdiff} \\
  q^{n+1} - \kappa \Delta t q_{xx}^{n+1} &= q^*, \label{eq:implicit-advdiff}
\end{align}
where $q^n$ is the value of $q$ at time step $n$, $q^*$ is an
intermediate value for $q$ and $\Delta t$ is the timestep.

We illustrate the use of FML for this example in three steps: 1)
define and label the weak form of the residual, 2) build solvers for
the explicit and implicit systems, 3) run the timestepping loop. Since
the intention of this example is to demonstrate FML rather than the
compatible finite element discretisation, we use the space of piecewise
quadratic continuous Lagrange polynomials, denoted by $P_2$, for the
finite element discretisation as this greatly simplifies the
forms. The finite element discretisation of equation
\ref{eq:1dadvdiff} is then
\begin{equation}\label{eq:weak-advdiff}
  \int_\Omega vq_t \diff x + c \int_\Omega v q_x \diff x + \kappa \int_\Omega v_x q_x \diff x = 0, \quad \forall v \in P_2,
\end{equation}
where $v\in P_2$ is the test function and the diffusion term has been
integrated by parts.

@@

\lstinputlisting[caption={This code defines the solver for the explicit step of the timestepping method i.e. equation \ref{eq:explicit-advdiff}. \texttt{qn} and \texttt{qstar} are Firedrake \texttt{Functions} with \texttt{qn} holding the values of $q$ at time step $n$ and \texttt{qstar} holding the values of $q$ after the explicit solve.}, label={lst:define-explicit}, language=python, firstline=40, lastline=55]{code/advection_diffusion_splitting_example.py}

The implicit solver is defined in a similar way (see listing
\ref{lst:define-implicit}). To create the left side of equation
\ref{eq:implicit-advdiff} we begin with the original residual, drop
the transport term (line 1), multiply all remaining terms that are not
the time derivative by $\Delta t$ (line 3) and replace the subject
with a trial function (line 5). The right side only contains the form
coming from the discretisation of the time derivative (line 7),
evaluated using the result of the explicit solve, \texttt{qstar} (line
9).

@@

\lstinputlisting[caption={This code is the timeloop implementing the scheme defined in equations \ref{eq:explicit-advdiff}-\ref{eq:implicit-advdiff} using the solvers defined in listings \ref{lst:define-explicit}-\ref{lst:define-implicit}.}, label={lst:timeloop}, language=python, firstline=73, lastline=79]{code/advection_diffusion_splitting_example.py}

This simple example demonstrates the power of manipulating the terms
in the equation based on the presence or absence of certain
labels. The full strength of FML becomes even more apparant when we
exploit information stored in the label value. We use this in two
important situations in Gusto. The first situation is when we require
a linearisation of the residual, for example when applying a Newton
method to the full nonlinear equations \JScomment{should we clarify
  that we mean a specific linearisation because Firedrake can do
  Newton methods without FML?}, or to test a new algorithm or model
setup on linear equations. The second situation is when we need to
evaluate source terms that come from physics schemes. We can
illustrate both situations with simple extensions to the above
example.

First consider the 1D viscous Burgers' equation
\begin{equation}
  u_t + uu_x + \nu u_{xx} = 0.
\end{equation}
